---
title: "Community, Mobility, and Vulnerability"
author: '[Katie Jolly](http://katiejolly.io)'
date: "January 14, 2018"
output:
  html_document:
    fig_height: 4
    fig_width: 6
    theme: cosmo
    toc: yes
    toc_float: yes
subtitle: "A look at Boston's 'insufficient/excessive heat' 311 reports"
---

***

```{r include = FALSE}
# packages

library(lubridate)
library(treemap)
library(highcharter)
library(viridisLite)
library(tidycensus)
library(leaflet)
library(tigris)
library(highcharter)
library(viridis)
library(sp)
library(raster)
library(rgdal)
library(rgeos)
library(maptools)
library(ggmap)
library(broom)
library(rvest)
library(RCurl)
library(RJSONIO)
library(tidyverse)

# data

boston_311 <- read_csv("311_cases_exercise.csv")

heat_311 <- boston_311 %>%
  filter(TYPE == "Heat - Excessive  Insufficient") %>%
  mutate(month = month(open_dt),
         year = year(open_dt),
         day = day(open_dt),
         mdy = mdy(paste(month, day, year, sep = " ")))

heat_days <- heat_311 %>%
  group_by(mdy) %>%
  summarize(total = n())
```

# Introduction

Between `r min(heat_311$mdy)` and `r max(heat_311$mdy)` there were `r nrow(heat_311)` reports made about insufficient/exessive heat in residential units. A few neighborhoods were affected disproportionately, as is the unfortunate truth of many disadvantages. Approximately 42% of the reports were made from the Dorchester and Roxbury neighborhoods. 

***

# Temporal patterns

Heat reports follow a seasonal trend, peaking in the summer and winter. However, there are many more calls made in the winter. My hypothesis for this is that someone with limited mobility would have a more difficult time going somewhere else, like a community center or other public space, when there are slick sidewalk conditions and frigid temperatures. Additionally, Boston's winter is more extreme than its summer, relatively speaking. 

```{r echo = FALSE}
hchart(name = "reports", heat_days, 
       "line", 
       hcaes(x = mdy, y = total)) %>%
  hc_yAxis(title = list(text = "Total reports"))
```

Reports spiked in late December/early January as a result of the far-below freezing temperatures that hit much of the United States, paricularly the east coast states. 


However, those spikes varied spatially across the city. Taking only the total values gives a skewed idea of how cold affects Boston's residents. To build off of the seasonal patterns, we can look at seasonal patterns by neighborhood. 

```{r echo = FALSE}
heat_311_neighborhood_date <- heat_311 %>%
  group_by(neighborhood, mdy) %>%
  summarize(total = n())

fntltp <- JS("function(){
  return this.point.mdy + ', ' +  this.series.yAxis.categories[this.point.y] + ':<br>' +
  Highcharts.numberFormat(this.point.value, 0) + ' report(s)';
}")


hchart(heat_311_neighborhood_date, "heatmap", hcaes(x = mdy, y = neighborhood, value = total)) %>% 
  hc_colorAxis(stops = color_stops(10, rev(magma(10))),
               type = "logarithmic") %>% 
  hc_yAxis(title = list(text = "    "), reversed = TRUE, offset = 20, tickLength = 0,
           gridLineWidth = 0, minorGridLineWidth = 0,
           labels = list(style = list(fontSize = "8px"))) %>% 
  hc_xAxis(title = list(text = "    ")) %>%
  hc_tooltip(formatter = fntltp)  %>% 
  hc_legend(layout = "vertical", verticalAlign = "top",
            align = "right", valueDecimals = 0) %>% 
  hc_size(height = 600)
```

Allston/Brighton, Dorchester, Greater Mattapan, Jamaica Plain, and Roxbury bear the vast majority of the burden for the heat-related requests. One intervention to take would be to advertise places where people can seek shelter during extreme weather in community meeting spaces, such as grocery stores or community center, that are specific to those areas. 

***

# Spatial patterns


```{r echo = FALSE, include = FALSE}
bos_inc <- get_acs(geography = "tract", 
              variables = c(medincome = "B19013_001"), 
              state = "MA",
              county = "Suffolk")

options(tigris_use_cache = TRUE)

bos_tract <- tracts(state = "MA", county = "Suffolk")

bos_joined <- geo_join(bos_tract, bos_inc, by = "GEOID") 

bos_joined <- subset(bos_joined, NAME.1 != "Census Tract 9901.01, Suffolk County, Massachusetts")
```


```{r include = FALSE, eval = FALSE}
community_centers_url <- read_html("https://www.boston.gov/community-centers")

centers <- community_centers_url %>%
  html_nodes(".cd-t") %>%
  html_text()

url <- function(address, return.call = "json", sensor = "false") {
 root <- "http://maps.google.com/maps/api/geocode/"
 u <- paste(root, return.call, "?address=", address, "&sensor=", sensor, sep = "")
 return(URLencode(u))
}
 
geoCode <- function(address,verbose=FALSE) {
 if(verbose) cat(address,"\n")
 u <- url(address)
 doc <- getURL(u)
 x <- fromJSON(doc,simplify = FALSE)
 if(x$status=="OK") {
 lat <- x$results[[1]]$geometry$location$lat
 lng <- x$results[[1]]$geometry$location$lng
 location_type <- x$results[[1]]$geometry$location_type
 formatted_address <- x$results[[1]]$formatted_address
 return(c(lat, lng, location_type, formatted_address))
 } else {
 return(c(NA,NA,NA, NA))
 }
}

names(centers) <- centers

community_centers_wide <- map_dfr(centers, function(x) geoCode(x))
  
community_centers <- community_centers_wide %>%
   rownames_to_column %>% 
   gather(var, value, -rowname) %>% 
   spread(rowname, value) %>%
  dplyr::select(-4)

colnames(community_centers) <- c("name", "latitude", "longitude", "address")

community_centers <- community_centers %>%
  mutate(latitude = as.numeric(latitude),
         longitude = as.numeric(longitude))
```

People with lower incomes are more likely to be renters, living in older buildings, or have less flexibility in their schedule, among other things. These factors all compound to make them more likely to be affected by cold winters. That is part of the reason why neighborhoods like Roxbury and Dorchester see disparate impacts during the winter. For simplicity (and time), I assumed a linear relationship between the median income in a census tract and the total number of heat-related reports in that tract. There was a statistically signficant, and I would argue practically significant as well, linear relationship between the two. In reality, I do not think a linear model would provide the best fit. Given more time, I would analyze different models to find the best one. I plotted the relationship below for visual reference. The model output is also printed below. 



```{r echo = FALSE, warning = FALSE, message = FALSE}
# reverse geocoding: lat long to census tract
heat <- SpatialPointsDataFrame(coords=heat_311[, c("Longitude", "Latitude")],
           data=heat_311[, c("neighborhood", "open_dt", "OnTime_Status", "CASE_STATUS", "LOCATION_STREET_NAME")],
           proj4string=CRS("+proj=longlat +datum=WGS84"))

heat <- spTransform(heat, "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0") # standardizing the projection

heat_tract <- over(x=heat, y=bos_joined) # aggregate

heat@data <- data.frame(heat@data, heat_tract) # combine data

heat_tracts <- heat@data # pull out only the data

# how many reports per census tract 
heat_tracts_calc <- heat_tracts %>%
  group_by(GEOID) %>%
  summarize(total = n(),
            estimate = mean(estimate),
            moe = mean(moe))


tract_lm <- lm(total~estimate, data = heat_tracts_calc) # model reports by income
broom_tract_lm <- tidy(tract_lm) # tidy it up

augmented_tracts <- augment(tract_lm) # calculate based on model

broom_tract_lm

highchart() %>%
  hc_add_series(name = "Observed data", heat_tracts_calc, "point", hcaes(x = estimate, y = total)) %>% # adding actual points
  hc_add_series(name = "Modeled data", augmented_tracts, "line", hcaes(x = estimate, y = .fitted)) %>% # adding fitted line 
  hc_tooltip(pointFormat = paste("{point.y} calls <br> Median income: ${point.x}"), crosshairs = TRUE) %>%
  hc_legend(enabled = TRUE) %>%
  hc_title(text = list("Modeling a linear relationship between heat complaints and income")) %>%
  hc_subtitle(text = list("calls ~ median income")) %>%
  hc_xAxis(text = list("Median income in census tract")) %>%
  hc_yAxis(text = list("Total number of reports in census tract")) %>%
  hc_credits(text = list("Source: Analyze Boston"))


```

Knowing that relationship, I've plotted the median values of the census tracts as a choropleth map. The red circles are locations that have reported `insufficient/excessive heat`. The BCYF icons are the geocoded locations of BCYF locations in Boston. This map starts to give an idea of how the spatial patterns play out on the ground. We can see that in general there are larger clusters of dots in lower income areas, which supports the original idea given by the model above. 

```{r echo = FALSE, message = FALSE, warning = FALSE}
community_centers <- read_csv("community_centers.csv")

heat_311 <- boston_311 %>%
  filter(TYPE == "Heat - Excessive  Insufficient")

pal <- colorBin("PuBu", domain = bos_inc$estimate, bins = 5)

comm_center_icon <- iconList(center = makeIcon("icons_bcyf.png", iconWidth = 22, iconHeight = 22))

leaflet(bos_joined) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(estimate),
    weight = 2,
  opacity = 1,
  color = "gray",
  dashArray = "1",
  fillOpacity = 0.9,
  highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = FALSE),
  label = bos_joined$NAME.1
  ) %>%
  addMarkers(data = community_centers,
             icon = comm_center_icon["center"],
             popup = community_centers$name,
             label = community_centers$name
) %>%
  addCircleMarkers(
    data = heat_311,
    radius = 11,
    color = "#af2a3c",
    stroke = FALSE, 
    fillOpacity = 0.8,
    popup = heat_311$LOCATION_STREET_NAME,
    label = heat_311$LOCATION_STREET_NAME,
    clusterOptions = markerClusterOptions()
  ) %>% addLegend(pal = pal, values = bos_joined$estimate, opacity = 0.7, title = "Median Income, 2016 ACS",
  position = "bottomright")

# find which community centers are in census tracts with certain percent under 60% of Mass median income and within 
```

***

# Recommendations 

But, there is one federal program in particular that can help. It is the Weatherization Assistance Program. This program gives grants to the states to help "weatherize" homes for low-income families that are overly-burdened by heating costs. In Boston the program is administered through Action for Boston Community Development. Unfortunately, President Trump has identified this program as one of the many assistance programs he wants to cut in order to save money[^2]. Putting that aside, in order to qualify, a family must earn less than 60% of the state-wide median income. 

[^2]:[This popular program helps keep homes warm in winter. Trump proposed eliminating it.](https://www.washingtonpost.com/news/powerpost/paloma/the-energy-202/2018/01/09/the-energy-202-this-popular-program-helps-keep-homes-warm-in-winter-trump-proposed-eliminating-it/5a53c16430fb0469e883ffad/?utm_term=.9602db11b3cb)

According to the 2016 5 year American Community Survey, Massachusetts had a state-wide median income of \$70,954 (+/- 343). Sixty percent of the state-wide median income is $42572. As a first-pass measure, I calculated which tracts have a median income equal to or less than \$42572. In those tracts, at least 50% of the households would qualify for weatherization funds from Action for Boston Community Development. In program year 2017, ABCD planned to spend \$538,150.00 weatherizing 90 units in Boston[^1]. Many home-improvement programs in the US target homeowners, but this federal program allows renters to apply with the permission of their landlord. I argue that we should work with local community centers (BCYF) to advertise the weatherization program. It would help target vulnerable families with young children as well as elderly people. Ideally, we could also work with places like libraries. I would include those if given more time. Since advertising campaigns can be resource-intensive, we should start with centers in tracts with a lower median income. I've highlighted those centers below. 

[^1]: [ABCD Annual Report](https://www.mass.gov/files/documents/2016/07/px/wapannualfile.pdf)

```{r echo = FALSE, message = FALSE, warning = FALSE}
low_income_tracts <- subset(bos_joined, estimate <= 42572)

center_clean <- community_centers %>%
  filter(!is.na(latitude))
center_sp <- SpatialPointsDataFrame(coords=center_clean[, c("longitude", "latitude")],
           data=center_clean[, c("name", "address", "latitude", "longitude")],
           proj4string=CRS("+proj=longlat +datum=WGS84"))

center_sp <- spTransform(center_sp, "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0") # standardizing the projection

center_tract <- over(x=center_sp, y=bos_joined) # aggregate

center_sp@data <- data.frame(center_sp@data, center_tract) # combine data

center_tract_data <- center_sp@data # pull out only the data

low_income_centers <- subset(center_tract_data, GEOID %in% low_income_tracts$GEOID)

leaflet(low_income_tracts) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = "yellow",
              dashArray = "1",
              weight = 1,
              color = "gray",
              fillOpacity = .8) %>%
  addMarkers(data = low_income_centers,
             label = ~name,
             popup = ~name,
             icon = comm_center_icon)
```

According to this analysis, coordinated communication efforts should start in these 12 centers. In the next round of analysis, I would perform a more sophisticated search using distance-based neighbors to identify patterns in surrounding tracts as well as look at centers near homes with prior reports. The 12 centers are Condon, Gallivan, Grove Hall Senior Center, Holland, Jackson/Mann, Madison Park, Mason Pool, Mattahunt, Perkins, Quincy, Shelburne, and Vine Street.

While WAP is certainly not the only heating assistance program or possible intervention, it is a good start and has similar criterion as the other programs. With further research I would identify alternatives and fine-tune the selection process for the best locations in which to advertise. 

The Low-Income Home Energy Assistance Program (LIHEAP) is another important program to advertise. It provides funds for heating costs, but is a more short-term solution. But in the winter, short-term is much preferable to the alternative: an unlivable home. 



